Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-04-04 _tex/paper2.tex
Built site for gh-pages

However, Thurstone designed Case V to provide a ``rather coarse
scaling'' of traits \citep[pp.~269]{Thurstone_1927b}, prioritizing
statistical simplicity over precise trait measurement. He cautioned that
its use ``should not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b}, as it imposes the most extensive set
of simplifying assumptions among all the proposed cases
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
(see Table~\ref{tbl-thurstone_cases}). Therefore, it is surprising that,
despite these limitations, CJ research has predominantly relied on Case
V to measure different traits, raising significant concerns about the
reliability and validity of such measurements in contexts where the
case's assumptions may not hold \citep[pp.~677]{Kelly_et_al_2022}.

Furthermore, since Thurstone's primary goal was to define a
psychological scale and ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not provide
specific guidance on how to use the measurement scores from Case V to
conduct hypothesis tests related to the measured traits. Although the CJ
practice has circumvented this issue by using the point estimates of the
scores or their transformations for such analyses, a fundamental
question remains: Is this method appropriate for addressing these
inquiries?

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, directly
influencing the outcome of pairwise comparisons. A thought experiment
can help illustrate this idea. In this experiment, researchers observe
the discriminal processes for two texts, similar to those shown in
Figure~\ref{fig-discriminal_process}. Additionally, the discriminal
dispersion for Text A remains constant, and the texts are uncorrelated
\((\rho=0)\). Figure~\ref{fig-dispersion} demonstrates that an increase
in the uncertainty associated with the perception of Text B relative to
Text A \((\sigma_{B} - \sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the probability area
where the discriminal difference distinctly favors Text B over Text A,
expressed as \(P(B>A)\), ultimately influencing the comparison outcome.
Additionally, the figure reveals that when the discriminal dispersions
of the texts are equal, as in the BTL model
\((\sigma_{B} - \sigma_{A}=0)\), the discriminal difference is more
narrow than when the dispersions differ. As a result, the discriminal
difference is more likely to favor Text B over Text A, as it is
represented by the shaded gray area.

@@

\subcaption{\label{fig-dispersion}Discrepancies in the dispersions of
stimuli}

@@

\subcaption{\label{fig-correlation}Correlation between stimuli}

@@

Notably, while the assumption of equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that the BTL
model may fail to capture the complexity of modern CJ data.
Specifically, the assumption of equal dispersions may not hold when
researchers assess complex traits or heterogeneous stimuli
\citetext{\citealp[pp.~376]{Thurstone_1927a}; \citealp[pp.~273]{Bramley_2008}; \citealp[pp.~678]{Kelly_et_al_2022}},
as these traits and stimuli can introduce judgment discrepancies due to
their unique characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide evidence of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.

These discrepancies may arise from two possible causes. First, the
discriminal processes of misfit texts may substantially overlap with
those of non-misfit texts. Second, misfit texts may represent outlying
observations with characteristics that make them difficult to compare
and position on the trait scale. Regarding the first cause, the overlap
can occur when misfit texts have equal dispersion and share the same
trait space as non-misfit texts or when they exhibit larger discriminal
dispersions. Although the BTL model can distinguish the first scenario,
it is inherently agnostic to the second.
Figure~\ref{fig-discriminal_process} illustrates how an increased
discriminal dispersion can widen the overlap between two texts, leading
to more judgment discrepancies. In the figure, Text B has a broader
discriminal dispersion than Text A, expanding the overlap between their
discriminal processes, a space where judges struggle to determine which
text possesses more of the trait.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\).

Finally, regarding the second cause, if researchers recognize that
misfit statistics may represent outlying observations, the conventional
CJ practice of excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
may unintentionally discard valuable information, introducing bias into
the trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}.
The direction and magnitude of these biases remain unpredictable, as
they depend on which stimuli researchers exclude from the analysis.

@@

As outlined in the previous section, researchers approach this process
in reverse. They begin by observing the outcomes and using the BTL model
to estimate the stimuli' discriminal differences and discriminal
processes. Given that the BTL model assumes independent discriminal
processes across comparisons, if this assumption holds in the data, the
model estimates a discriminal difference distribution that accurately
reflects the \emph{true} discriminal difference between the texts. Once
more, the accuracy of the discriminal difference ensures reliable
estimates for the discriminal processes of the texts.

@@

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on various intricate characteristics of the
stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, when assessed, are unlikely to be
equally weighted or opposing. As a result, they may exert an uneven
influence on judges' perceptions, creating biases that resist
cancellation. For example, this could occur when a judge assessing the
argumentative quality of a text places more weight on its grammatical
accuracy than other judges, thereby favoring texts with fewer errors but
weaker arguments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases could introduce dependencies between the
stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}. While direct
evidence for this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the notion that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli, a statistical phenomenon known as clustering
\citep{Everitt_et_al_2010}. Although the CJ literature acknowledges the
existence of such hierarchical structures, the statistical approaches to
account for this additional source of dependence have been insufficient.
For instance, when CJ data incorporates multiple samples of stimuli from
the same individuals, researchers frequently rely on (averaged) point
estimates of the BTL scores to conduct subsequent analyses and tests at
the individual level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.

@@

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what we refer to as the
\emph{conceptual-population model}. This model assumes that researchers
have access to a \emph{conceptual population} of comparative data, that
is, data representing all repeated judgments made by every available
judge for each pair of stimuli produced by each pair of individuals in
the population. Conversely, Section~\ref{sec-theory-theoretical_SC}
integrates the assessment design features into what we refer to as the
\emph{sample-comparison model}. This model assumes a more realistic
scenario where researchers only have access to a sample of judges,
individuals, stimuli, and comparisons from the conceptual population.

@@

In the conceptual-population model, we assume an idealized scenario
where researchers have access to a \emph{conceptual population} of
comparative data --data representing \emph{all repeated judgments made
by each judge for every stimulus pair produced by each pair of
individuals} in the population. This assumption allows us to integrate
Thurstone's theoretical principles and propose innovations to address
some of the issues raised in Section~\ref{sec-theory-issues}.

@@

\begin{equation}\phantomsection\label{eq-mat11}{
I = \begin{bmatrix}
1 \\
2 \\
\vdots \\
h \\
\vdots \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A}-1 \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K}-1 \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{J} & n_{K} \\
\end{bmatrix}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mat12}{
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K}-1 \\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & i & a & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K}-1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
\end{bmatrix}
}\end{equation}

@@

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{Here, \emph{relevant} refers
  to variables that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. These topics are beyond the scope of this study,
  thus, readers seeking a more profound understanding can refer to
  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, as well as
  introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020},
  and \citet{McElreath_2020} useful. For more advanced study, seminal
  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
  \citet{Hernan_et_al_2020}, are recommended.} student-related variables
\((X_{I})\), such as the teaching method, causally influence the latent
variable representing students' writing-quality trait \((T_{I})\). It
also shows how \(T_{I}\) and \emph{relevant} \footnote{refer to footnote
  1.} text-related variables \((X_{IA})\) (e.g., text length) causally
influence the texts' written-quality trait \((T_{IA})\), the first
primary latent variable of interest. Additionally, the branch accounts
for idiosyncratic errors: \(e_{I}\) captures variations in students'
traits unexplained by \(X_{I}\), while \(e_{IA}\) captures variations in
texts' traits unexplained by \(T_{I}\) or \(X_{IA}\). Here, \(X_{I}\) is
an observed matrix with \(n_{I}\) rows and \(q_{I}\) columns
(variables), and both \(e_{I}\) and \(T_{I}\) are latent vectors of
length \(n_{I}\). Similarly, \(X_{IA}\) is an observed matrix with
dimension \(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns (variables),
while \(e_{IA}\) and \(T_{IA}\) are latent vectors of length
\(n_{I} \cdot n_{A}\).

@@

\textbf{CRediT authorship contribution statement:}
\emph{Conceptualization:} S.G., S.DM., T.vD., and J.M.R.E;
\emph{Methodology:} S.DM., T.vD., and J.M.R.E; \emph{Software:}
J.M.R.E.; \emph{Validation:} J.M.R.E.; \emph{Formal Analysis:} J.M.R.E.;
\emph{Investigation:} J.M.R.E; \emph{Resources:} S.G., S.DM., and T.vD.;
\emph{Data curation:} J.M.R.E.; \emph{Writing - original draft:}
J.M.R.E.; \emph{Writing - review and editing:} S.G., S.DM., and T.vD.;
\emph{Visualization:} J.M.R.E.; \emph{Supervision:} S.G. and S.DM.;
\emph{Project administration:} S.G. and S.DM.; \emph{Funding
acquisition:} S.G. and S.DM.

>>>>

However, Thurstone developed Case V prioritizing statistical simplicity
over precise trait measurement, and without providing clear guidance on
how to use its measurement scores for hypothesis testing. Specifically,
he cautioned that its use ``should not be made without (an) experimental
test'' \citep[pp.~270]{Thurstone_1927b}, as the case imposes the most
extensive set of simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
(see Table~\ref{tbl-thurstone_cases}). Moreover, because Thurstone's
primary goal was to provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b} and ``allocate the compared stimuli on
this continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not address
how to use the resulting measurement scores for hypothesis testing.
Thus, given these limitations, it is surprising that CJ research has
predominantly relied on Case V to measure different traits, raising
significant concerns about the reliability and validity of such
measurements in contexts where the case's assumptions may not hold
\citep[pp.~677]{Kelly_et_al_2022}. Furthermore, although the CJ practice
has attempted to address the issue of hypothesis testing by using the
scores' point estimates or their transformations, a critical question
remains: Is this approach suitable for addressing these inquiries?

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, directly
influencing the outcome of pairwise comparisons. A thought experiment
can help illustrate this idea. In it, researchers observe the
discriminal processes for two texts, A and B, assuming that the
dispersion for Text A remains constant and that the two texts are
uncorrelated \((\rho=0)\). Figure~\ref{fig-dispersion} demonstrates that
an increase in the uncertainty associated with the perception of Text B
relative to Text A \((\sigma_{B} - \sigma_{A})\), broadens the
distribution of their discriminal difference. This broadening affects
the probability area where the discriminal difference distinctly favors
Text B over Text A, expressed as \(P(B>A)\), ultimately influencing the
comparison outcome. Additionally, the figure reveals that when the
discriminal dispersions of the texts are equal, as in the BTL model
\((\sigma_{B} - \sigma_{A}=0)\), the discriminal difference distribution
is more narrow than when the dispersions differ. As a result, the
discriminal difference is more likely to favor Text B over Text A, as it
is represented by the shaded gray area.

@@

\subcaption{\label{fig-dispersion}Discriminal Difference distribution
under varying discrepancies in stimuli dispersions}

@@

\subcaption{\label{fig-correlation}Discriminal Difference distribution
under varying levels of correlation between stimuli}

@@

Notably, while assuming equal dispersions simplifies the trait
measurement model, evidence from the CJ literature suggests that this
assumption may fail to capture the complexity of modern CJ data. In
particular, the assumption may not hold when researchers assess complex
traits or heterogeneous stimuli
\citetext{\citealp[pp.~376]{Thurstone_1927a}; \citealp[pp.~273]{Bramley_2008}; \citealp[pp.~678]{Kelly_et_al_2022}},
as these traits and stimuli can introduce judgment discrepancies due to
their unique characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide indications of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.
These misfit texts may exhibit larger discriminal dispersions or
represent outliers --texts with distinctive characteristics that deviate
markedly from the rest of the sample \citep{Grubbs_1969}. In either
case, comparing misfit texts results in more judgment discrepancies, yet
the BTL model neither accounts for these cases nor offers any means to
address them.

Significant statistical and measurement issues can arise when the
assumption of equal dispersions between stimuli does not hold.
Specifically, the BTL model may overestimate the trait's reliability,
that is, the degree to which the outcome accurately reflects the
\emph{true} discriminal differences between stimuli. This
overestimation, in turn, results in spurious conclusions about these
differences \citep[pp.~370]{McElreath_2020} and, by extension, about the
underlying discriminal processes of stimuli. Figure~\ref{fig-dispersion}
also illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Furthermore, if researchers
acknowledge that misfit statistics may represent outlying observations,
the common CJ practice of excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
may unintentionally discard valuable information, introducing bias into
the trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}.
The direction and magnitude of these biases remain unpredictable, as
they depend on which stimuli researchers exclude from the analysis.

@@

@@

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on various intricate characteristics of the
stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, when assessed, are unlikely to be
equally weighted or opposing. As a result, they may exert an uneven
influence on judges' perceptions, creating biases that resist
cancellation. For example, this could occur when a judge assessing the
argumentative quality of a text places more weight on its grammatical
accuracy than other judges, thereby favoring texts with fewer errors but
weaker arguments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases could introduce dependencies between the
stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}. While direct
evidence for this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} and \citet{vanDaal_et_al_2016} demonstrate
the presence of such biases, supporting the notion that the factors
influencing pairwise comparisons may not always cancel out.

In the second scenario, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli, a statistical phenomenon known as clustering
\citep[pp.~86]{Everitt_et_al_2010}. For instance, when multiple texts
are produced by the same individual, these texts are likely to share
common features, such as writing style or general quality. Although the
CJ literature acknowledges the existence of such hierarchical
structures, the statistical approaches to account for this additional
source of dependence have been insufficient. For instance, when CJ data
incorporates multiple samples of stimuli from the same individuals,
researchers frequently rely on (averaged) point estimates of the BTL
scores to conduct subsequent analyses and tests at the individual level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.

@@

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what we refer to as the
\emph{conceptual-population model}. This model assumes an idealized
scenario where researchers have access to a \emph{conceptual population}
of comparative data, that is, data representing all repeated judgments
made by every available judge for each pair of stimuli produced by each
pair of individuals in the population. Conversely,
Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
features into what we refer to as the \emph{sample-comparison model}.
This model assumes a more realistic scenario where researchers only have
access to a sample of judges, individuals, stimuli, and comparisons from
the conceptual population.

@@

In the conceptual-population model, the idealized scenario of a
\emph{conceptual population} of comparative data enables the integration
of Thurstone's theoretical principles and provides a foundation for
proposing innovations aimed at addressing some of the issues discussed
in Section~\ref{sec-theory-issues}.

@@

\begin{equation}\phantomsection\label{eq-mat11}{
I = \begin{bmatrix}
1 \\
2 \\
\vdots \\
h \\
i \\
\vdots \\
n_{I}
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
\vdots \\
j \\
\vdots \\
n_{J}
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{A} \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & 1 \\
\vdots & \vdots \\
n_{I} & n_{A} \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
1 & n_{K} \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & 1 \\
\vdots & \vdots \\
n_{J} & n_{K} \\
\end{bmatrix}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mat12}{
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 1 & 2 & 1 & n_{K} \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
h & i & a & b & j & k \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 1 \\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & 2 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
n_{I} & n_{I} & n_{A}-1 & n_{A} & n_{J} & n_{K} \\
\end{bmatrix}
}\end{equation}

@@

The top branch of DAG -~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{Here, \emph{relevant} refers
  to variables that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. These topics are beyond the scope of this study,
  thus, readers seeking a more profound understanding can refer to
  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, as well as
  introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020},
  and \citet{McElreath_2020} useful. For more advanced study, seminal
  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014}, and
  \citet{Hernan_et_al_2020}, are recommended.} student-related variables
\((X_{I})\), such as teaching method, and students' idiosyncratic errors
\(e_{I}\) causally influence the latent variable representing students'
writing-quality trait \((T_{I})\). The error term \(e_{I}\) captures
variations in students' traits unexplained by \(X_{I}\). This branch
also shows how \(T_{I}\), along with \emph{relevant} \footnote{refer to
  footnote 1.} text-related variables \((X_{IA})\) (e.g., text length),
and texts' idiosyncratic errors \(e_{IA}\) causally influence the texts'
writing-quality trait \((T_{IA})\), the first primary latent variable of
interest. The error term \(e_{IA}\) captures variations in the texts'
traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). In this
model, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(q_{I}\)
columns (variables), and both \(e_{I}\) and \(T_{I}\) are latent vectors
of length \(n_{I}\). Similarly, \(X_{IA}\) is an observed matrix with
dimensions \(n_{I} \cdot n_{A}\) rows and \(q_{IA}\) columns
(variables), while \(e_{IA}\) and \(T_{IA}\) are latent vectors of
length \(n_{I} \cdot n_{A}\).

@@

\textbf{CRediT authorship contribution statement:}
\emph{Conceptualization:} J.M.R.E, T.vD., S.DM., and S.G.;
\emph{Methodology:} J.M.R.E, T.vD., and S.DM.; \emph{Software:}
J.M.R.E.; \emph{Validation:} J.M.R.E.; \emph{Formal Analysis:} J.M.R.E.;
\emph{Investigation:} J.M.R.E; \emph{Resources:} T.vD. and S.DM.;
\emph{Data curation:} J.M.R.E.; \emph{Writing - original draft:}
J.M.R.E.; \emph{Writing - review and editing:} T.vD., S.DM., and S.G.;
\emph{Visualization:} J.M.R.E.; \emph{Supervision:} S.G. and S.DM.;
\emph{Project administration:} S.G. and S.DM.; \emph{Funding
acquisition:} S.G. and S.DM.
